\documentclass[defaultstyle, allheadings, 11pt]{thesis}

\usepackage{amssymb}		% to get all AMS symbols
\usepackage{graphicx, booktabs}		% to insert figures
\usepackage{hyperref}		% PDF hyperreferences??
\usepackage{xcolor, soul}

\usepackage{framed}
\usepackage{caption, subfig}
\captionsetup[figure]{labelfont=bf, font=small}

\usepackage{fontspec, titlesec}
\setmainfont{Noto Serif}
\setsansfont{Noto Sans}
\setmonofont{Noto Sans Mono}

\titleformat{\chapter}
[block] % shape
{\sffamily\LARGE\bfseries} % format
{\thechapter.} %label formatting
{0em}{}

\titleformat{\section}[hang]
{\sffamily\Large\bfseries}
{\thesection}{1em}{}

\titleformat{\subsection}[hang]
{\sffamily\large\bfseries}
{\thesubsection}{1em}{}

\titleformat{\subsubsection}[hang]
{\normalfont\bfseries}
{}{0em}{}


\input macros.tex


%%%%%%%%%%%%   All the preamble material:   %%%%%%%%%%%%

\title{Retooling E-Textiles for Coproduction: \revision{Weaving Circuitry as Cloth}}

\author{Shanel}{Wu}

\otherdegrees{B.S., Harvey Mudd College, 2016}

\degree{Doctor of Philosophy}		%  #1 {long descr.}
	{Ph.D., Creative Technology Design}		%  #2 {short descr.}

\dept{ATLAS Institute}			%  #1 {designation}
	{}		%  #2 {name}

\advisor{Prof.}				%  #1 {title}
	{Laura Devendorf}			%  #2 {name}

\reader{Mirela Alistar}		%  2nd person to sign thesis
\readerThree{Leah Buechley}		%  3rd person to sign thesis
\readerFour{Daniel Leithinger}
\readerFive{Gregory Whiting}


\abstract{  \OnePageChapter	% because it is very short

E-textiles (sometimes called ``smart textiles'') is an emergent technology that integrates electronics with textile materials and structures. While the field promises a future industry with applications for personalized wearables, healthcare, smart vehicles, and more, there are still very few viable products for mainstream consumers. My dissertation focuses on concerns for the sustainability of future e-textiles, and explores the design opportunities presented by craft that engage with such issues. Through the lens of \keyterm{craft}, I will develop \keyterm{retooling for coproduction} as one possible way for e-textiles designers to practice and explore sustainability through hands-on making. I begin with some background on treatments of craft and sustainability in human-computer interaction (HCI) research and adjacent communities, focusing on digital fabrication, textiles, and social justice. With this ``design orientation", I will then present the research undertaken during my PhD studies as case studies in crafting e-textiles design tools that resulted in deeper understandings of textile practices and histories. These retools include generative tactics, woven structures, and tangible computational interfaces, motivating craft's relevance to \citedterm{retooling}---how technological tools can support social justice efforts; and \citedterm{coproduction}---an awareness of how technology and society mutually shape one another; in reference to design justice and feminist science, technology, and society (STS) studies. Lastly, I will present speculations for further sustainable e-textiles retooling and the future research directions in which my craft may take me.

% E-textiles (also referred to as ``smart textiles") is an emergent technology that integrates electronics with textile materials and structures. While the field promises a future industry with applications for personalized wearables, healthcare, smart vehicles, and more, there are still very few viable products for mainstream consumers. With concerns about durability, usability, and manufacturability along with open problems surrounding the sustainability, inclusivity, and accessibility of future e-textiles, my dissertation argues that the e-textiles field will greatly benefit from prioritizing \keyterm{craft} as a central value. Through the lens of craft, I will develop \keyterm{retooling for coproduction} as one possible way for e-textiles designers to orient the technical aspects of their practices towards broader sociopolitical discourses, such as sustainability. This dissertation begins with some background on craft's role in related human-computer interaction (HCI) research, and in the adjacent domains of textiles and social justice. With this premise as my ``design orientation", I will then present the research undertaken during my PhD studies as case studies in crafting e-textiles design tools that resulted in deeper understandings of textile practices and histories. These retools include design tactics, woven structures, and tangible computational interfaces, 
% motivating craft's relevance to \citedterm{retooling}---how technological tools can support social justice efforts; and \citedterm{coproduction}---an awareness of how technology and society mutually shape one another; concepts that reference design justice and feminist science, technology, and society (STS) studies. Lastly, I will present speculations for further sustainable e-textiles retooling and the future research directions in which my craft may take me.
%will be developing its own tools, or \textit{retooling}, distinct from those of either textiles or electronics. This dissertation, I will review smart textiles research in human-computer interaction (HCI), design, and other relevant domains to motivate the importance of retooling in an emergent design domain, as well as motivating the concept of coproduction as an alternative to discourses that characterize smart textiles integration as a "hybrid" or "composite" system composed of some parts from textiles and others from electronics. 
%I will then present the completed research that I have undertaken during my PhD studies, which focuses on woven smart textiles, to demonstrate how a design orientation of retooling smart textiles for coproduction can produce deeper understandings of craft practices and textiles histories, and thus generate new design opportunities for future smart textiles technologies. Furthermore, by endorsing the justice-oriented and feminist intentions of the conceptual origins of both "retooling" and "coproduction" in design, I also align my focus on technological design tools with values-driven design discourses around sustainability and intersectional justice. Finally, I will propose the final component of my dissertation needed to complete my argument, which will be developing an embedded hardware/software design tool for smart textiles weaving interfaces to embody both retooling and coproduction.

%	Often the abstract will be long enough to require
%	more than one page, in which case the macro
%	``$\backslash$OnePageChapter'' should {\it not}
%	be used.

%	But this one isn't, so it should.
	}

\dedication[Dedication]{	% NEVER use \OnePageChapter here.
	To my family,

	by blood and by choice,

	furred and feathered,

	in this life and other lifetimes
}

\acknowledgements{	\OnePageChapter	% *MUST* BE ONLY ONE PAGE!
	Thank you to my labmates in the Unstable Design Lab for encouraging me to pursue this research path, and for building a nurturing work environment: Mikhaila, Xavier, Emma, Jolie, Deanna, Eldy, Jordan, Lily, Sasha, Nikita, Etta, Chad, and Steven.
	
	Thank you to my advisor, Laura Devendorf, for her academic and professional guidance, but also for her patience, emotional support, and willingness to learn with me about my ADHD.

	Thank you to my roommates, who helped me build a home in Boulder --- especially Monse.

	Thank you to my backyard flock, my babies, for teaching me how to live with other species. And the nutritious eggs. Pepita, Conchita, Panther, Jaguar, Lupita, Canelia, and \textit{in memoriam:} Jefita and Griselda.

	Thank you to the /r/ADHD Discord server and my co-conspirators in AcaDHD.

	Thank you to Miranda and Ben --- love you lots.

	Thank you to the land and this planet for giving me time and space.

	And finally, thank you to all who enable my messy, weird, crafty existence.


	}

% \IRBprotocol{E927F29.001X}	% optional!

\ToCisShort	% use this only for 1-page Table of Contents

\LoFisShort	% use this only for 1-page Table of Figures
%\emptyLoF	% use this if there is no List of Figures

% \LoTisShort	% use this only for 1-page Table of Tables
\emptyLoT	% use this if there is no List of Tables

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%       BEGIN DOCUMENT...         %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\setlength{\parindent}{0.4in}
\raggedright

\input ch_1_intro.tex
\input ch_2_background.tex
\input ch_3_e-textiles.tex
\input ch_4_unfabricate.tex
\input ch_5_loom-pedals.tex
\input ch_6_speculations.tex
\input ch_7_conclusion.tex

%%%%%%%%%   then the Bibliography, if any   %%%%%%%%%
\bibliographystyle{acm}	% or "siam", or "alpha", etc.	% list all refs in database, cited or not
\bibliography{refs}		% Bib database in "refs.bib"

%%%%%%%%%   then the Appendices, if any   %%%%%%%%%
% \appendix
% \input apx_glossary.tex


\end{document}

