\chapter{\revision{Epilogue: To a Future Researcher}}

\noindent\revision{Dear friend,}

\revision{Academia has been a strange and difficult place to pursue my love of craft. I have had more freedom than any other point in my life to research what I felt was most important and build something according to my own vision. My rent for this space is paid in words. Sometimes, it's been \textit{really} hard to pay this rent by getting a manuscript submitted or this dissertation finished. So my first piece of advice is to really reflect on your relationship with writing, and invest energy in your personal writing toolkit.}

\revision{Secondly, I would generally reflect on yourself, for research purposes. We do not pretend to be objective creators who force materials into our designs. You are bringing your own unique perspective and subjectivity to the research, which colors your interactions with the materials, your tools, and your colleagues and collaborators. By reflecting on how your identities and experiences are part of your design practice, you also add this history to your pool of design resources.}

\revision{Sometimes, the wickedness of our current sustainability problems would send me into a spiral of climate anxiety and despair. If you find yourself in a similar position when researching sustainability in design, take a break. Go warp a loom or untangle some yarn. Do something that brings you joy and satisfaction.}

\revision{But with all of this introspection, reading, writing, tinkering, and making, don't spend too much time alone. Academic research lends itself to many hours working by yourself, and when your work involves fabrication, developing hardware and software, and testing things in a lab, it's even easier to just burrow into your own world. By definition, doing a PhD, especially in such an interdisciplinary space, is about doing something that nobody else has ever done. This can be extremely isolating, because nobody can perfectly empathize with all of your struggles.} 

\revision{I won't pretend that these pieces of advice are guaranteed to work, but I do want them to serve as reminders that you have a community. You have me and other people on our own research paths, that all wander off individually, but we're all still within shouting distance. I wish you the best of luck, and lots of strength, curiosity, and love.}
% There have been many points where I almost couldn't come up with the words to make payment in the form of a research paper or even this dissertation. 
